Estatus social subjetivo, posición objetiva y percepción de meritocracia: una aproximación empírica al caso de Chile

### 2023-01-06 paper-blind.Rmd
reduce words

---
 title: "Status, Class, Subjective status and meritocracy perception"
 # author: Julio Iturra ^[Bremen International Graduate School of Social Sciences (jiiturra@uni-bremen.de)]
 bibliography: "input/bib/ess-merit.bib"
 csl: "input/bib/apa6.csl"
 css: "input/css/custom.css"
 output:
     bookdown::html_document2:
         number_sections: true
     bookdown::pdf_document2:
         template: default
         toc: false
         number_sections: false
     bookdown::word_document2:
         reference_docx: template.docx
 geometry: margin=1in
 header-includes:
   # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel} # para decimales con commas en vez de punto
   - \usepackage{times}           # Times New Roman
   - \usepackage{caption}
   - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
   - \usepackage{graphicx}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
 linestretch: '1.5'
 fontsize: 12pt
 link-citations: yes
 linkcolor: blue
 editor_options:
   chunk_output_type: inline
 abstract: "What is the role of status, class and subjective status on perceived meritocracy? To what extent can subjective status affect the objective experiences given by income, education and social class on meritocracy perception? To answer these questions, we use nationally representative data from the 2018 wave of the Longitudinal Social Survey of Chile - ELSOC (N = 2,652). Perceived meritocracy constitutes the descriptive views about how effort and talent are rewarded in society. Contrary to our expectations, income and social class do not influence perceived meritocracy. However, in line with the hypothesis, individuals with higher educational attainment hold critical views about meritocracy, and those who allocate themselves close to the top groups in society perceive higher rewards regarding effort and talent. We also found that subjective status increases the meritocracy perception of those groups on top of the social hierarchy concerning income, education and social class."  
---

```{r render,include=FALSE, eval=FALSE}
 rmarkdown::render("paper-eng.Rmd")
```

@@

During the last decades, the rise of economic inequality [@piketty_capital_2014] and the stagnation of social mobility [@oecd_broken_2018] has increased the attention of social scientists on how societies perceive the functioning of meritocratic rules - that efforts and talents are rewarded - and its role in economic inequality justification [@mijs_meritocracy_2020]. Meritocracy has been related to the justification of social inequalities, as individual achievements and location in the class structure are mostly attributed to individual agency rather than structural constraints [@frei_fruto_2018]. In this regard, the literature  has suggested that individuals' views about the labour market create an image of labour relations and how economic resources are distributed [@Svallfors2006]. Furthermore, empirical evidence has shown that belonging to non-manual and professional social classes is negatively associated with perceived meritocracy, which is more salient in societies with higher levels of economic inequality and low social class mobility [@Canales2015lse].

Social stratification can be studied from different approaches. On the one hand, studies on objective social stratification have analysed how objective social inequality is expressed in  economic and cultural resources [@Kerbo2003], usually associated with the concept of social class. On the other hand, @Kluegel-Smith1986 point out that studying social inequality from a subjective standpoint takes into account elements of individual experience in the class structure. The concepts of class and status have been crucial for studying social inequality from the social sciences [@Ridgeway2014]. Whereas class analysis mainly focuses on occupations, structural position in the labour market and levels of authority and control over work [@Erikson1992], status is understood as a structure of relationships perceived as superior, equitable, or inferior among individuals [@Chan2007].

In sociology, the concept of class is proximate to occupational class schemes, but also it has been linked to variables such as income and education. Regarding subjective aspects, although there are scales of subjective social class [@Elbert2018], research on subjective social status has been developed more [@zhou_trajectory_2021]. In this respect, it has been shown that the perception of an individual or group's standing in the social hierarchy compared to a reference group, such as society [@Evans2004] or their local communities [@condon_economic_2020] can also explain perceptions about meritocracy [@Castillo2018]. It has been stated that individuals are likely to compare themselves with people or groups perceived as similar, contributing to diminishing cognitive dissonance and seeking to assemble better estimations about their societal position [@Merton1968;@festinger1954theory]. Whereas empirical evidence has shown a persistent middle-class bias toward the center of the social hierarchy, where high-status individuals underestimate their position while low-status individuals overestimate it [@Cruces2011]. At this point, the literature has suggested that the characteristics of the closest reference groups shape subjective estimations, in which images of society are influenced by the homophily of their close networks [@Evans2017 ]. Thus, higher subjective social status is expected to be associated with higher perceived meritocracy and also positively moderate the influence of class and status [@hvidberg_social_2020].

This study aims to understand how objective class and status influence meritocracy
perception. In addition, it discusses the role of subjective social status and how it interacts with social class, education and income.

@@

First, _merit_ is understood as the combination of an individual's effort and ability [@Sen2000]. Second, the meritocratic ideal assumes that status achievement is not constrained by the distribution of opportunities in society [@mccall_exposure_2017]. Both constitute the substantive core of how meritocracy can be understood. Nevertheless,  studies have paid little attention to the construct measurement of concepts related to meritocracy, and low empirical precision is often noticed due to the indistinct use of descriptive and normative dimensions [@Castilloetal2019].

For instance, studies on meritocracy have made undifferentiated use of concepts such as *Meritocratic Beliefs* [@Kunovich2007; @McCoy2007; @Ellis2017; @mijs_how_2021a; @bernardo_meritocracy_2021], *Meritocratic Ideals*, [@Reynolds2014] or *Perceived Meritocracy* [@Xian2017], which constitutes a difficulty to empirically assess the role of social status on the perception of meritocracy. This lack of distinction in the literature suggests considering a more accurate definition of the construct. Furthermore, this would lead to more precise inferences and better evidence regarding the empirical study of the subjective aspects of meritocracy [@Castilloetal2019].

Regarding the perception and preferences distinction, @Aalberg2003 defines _perception_ as the subjective estimation of an external object depicted by individuals. Therefore, the descriptive (*how is*) dimension does not consider the normative evaluation of the object (*should be*). On the other hand, @druckman_preference_2000 define the normative dimension as the expression of preferences, understood as evaluating an external object based on previous experience and information. In addition, @Feldman2001 suggests that preferences are also driven by the structure and internalisation of norms and values [@Kulin2013]. On the other side, when we talk about meritocracy perception the descriptive dimension is addressed, referring to "how meritocracy is" rather than "how it should be". In this line, @Garcia2001 posits that meritocracy perception corresponds to the degree to which individuals consider that their society complies with the principles of a meritocracy. In other words, how individuals perceive to what extent society allocates rewards (e.g. wages or jobs) based on effort and talent.

@@

### Status 
The evidence on the association of social status with meritocracy follows the general assumption that individuals with higher social status have a higher meritocracy perception. From a self-interest approach [@Meltzer81] it is possible to argue that individuals who belong to the higher strata of the income distribution have a favourable perception of meritocracy to the extent that they have accumulated economic resources as a result of their effort and talent. In this line, empirical evidence suggests that higher status positions are related to lower inequality perception [@Evans2017] and the highest justification of economic inequality [@Trump2017; @Jasso2015]. Regarding meritocracy, the core assumption is that merit explains status achievement. In this sense, if higher perceived equity is related to structural opportunities to get ahead, it is reasonable to argue that meritocracy constitutes the central mechanism for justifying differences in social status [@McCoy2007].

Evidence suggests a positive relationship between income and perceptions of meritocracy regarding the importance of individual merit in getting ahead in life [@Reynolds2014; @Xian2017], as well as what extent individuals perceive intelligence and skills are rewarded [@Castilloetal2019; @Duru-bellat2012]. Nevertheless, there is mixed evidence regarding how lower-income individuals rationalise inequality and meritocracy. On the one hand, @Newman2015 shows proof in favour of the *activated conflict theory* demonstrating the poor hold lower beliefs in meritocracy in unequal localities, arguing economically heterogeneous contexts induce social comparisons leading to higher awareness about social status. On the other hand, following the *relative power theory*, @Solt2016 challenged the previous evidence showing the poor in unequal contexts to strongly support meritocracy. The authors suggest the rich hold higher power due to income and wealth concentration, allowing them to disseminate legitimising ideas about economic inequality. In this context, the poor are less critical towards dominant beliefs like meritocracy, but it also serves as a psychological mechanism for resilience deprivation [@Solt2016,p.4]. On top of this debate, @morris_paradox_2022 findings show firmer meritocracy beliefs among low-income individuals in highly unequal contexts, but the differences tend to fade as inequality declines.

Based on a material self-interest approach is possible to argue that individuals with higher economic resources attribute more importance to merit and therefore perceive rewards in society are distributed according to it, which leads us to establish:

> $H_{\text{1a}}$: the position in the income distribution is positively associated with perceived meritocracy.

In the same way, income reflects an individual's economic resources, the study of meritocracy concerning educational attainment has been mainly addressed by studies on social stratification [@Goldthorpe2003]. Studies on meritocracy perception have discussed two theoretical approaches. On the one hand, the *reproductionist* approach [@Bourdieu2009; @bernstein_class_2003] supports the *socialisation* hypothesis. The argument suggests school socialisation operates as a normative reinforcement of meritocratic ideals, then educated individuals are more willing to support merit as the main mechanism for legitimising inequality [@Lampert2013]. On the other hand, the *instruction* hypothesis suggests that perceived meritocracy tend to weaken in individuals with higher educational attainment because they would be more aware of the influence of structural factors on status achievement [@Duru-bellat2012; @Schneider2015] taking into account educational mobility expriences [@Canales2015lse] and the increasing chances to engage with more complex political ideas [@bubak_perceptions_2019].

Empirical evidence provides support for both approaches. However, it should be noted that these results can be attributed to differences in the measurement of perceived meritocracy. On the one hand, studies report evidence in favour of the socialisation hypothesis have pointed out that greater educational attainment is associated with greater attributed importance of individual factors in getting ahead in life [@Xian2017; @Reynolds2014; @mijs_paradox_2021], as well as determining the attainment of a better salary [@Kunovich2007]. On the other hand, evidence in favour of the instruction hypothesis shows that individuals with higher educational attainment strongly disagree that people are rewarded for their effort and talent [@Duru-bellat2012; @Castilloetal2019]. Additionally, the argument of the instructional hypothesis can be complemented by the fact that individuals with greater awareness of inequality may experience feelings of injustice and relative deprivation since individuals with the same educational attainment have access to unequal rewards [@Adams1965; @Smith-eta2020_relative-deprivation]. In this sense, it has been shown that individuals with higher educational attainment are those who perceive greater wage inequality [@Castillo2012a_justice]. If perceived meritocracy is understood as " the perception society is based on the principles of meritocracy" [@Garcia2001], it corresponds to the descriptive appraisals of the link between merit and rewards. Then, we can argue that:

> $H_{\text{1b}}$: educational attainment is negatively associated with perceived meritocracy.

@@

The relationship between structural position and meritocracy has mainly focused on income and educational attainment, setting social class aside from the discussion. However, it has been shown that occupation represents a characteristic that synthesises educational attainment and income level [@Ganzeboom1992], therefore is considered one of the main status symbols in contemporary societies [@Chan2004; @Chan2007]. Furthermore, studies on social mobility have shown the relevance of studying social inequality through class schemes because they take into account the occupation, autonomy and authority in the workspace to classify individuals, which comprise a substantive distinction [@Erikson1992; @Erikson2002]. In this sense, evidence on social stratification suggests that occupation-based class schemes assume that individuals are similar in terms of their structural position in the labour market and levels of authority and control over work [@Connelly2016]. However, @Chan2004 argue that common characteristics do not preclude the possibility of heterogeneity within classes, for instance, in terms of cultural consumption behaviour and political preferences [@Chan2007].

The literature on distributive preferences has suggested social class reflects individuals' experience in the labour market regarding job opportunities and wages [@Svallfors2006]. In a similar vein, @Kulin2013 demonstrated lower classes being exposed to higher labour market risks, such as the likelihood of job loss and lower wages, led individuals in material deprivation and risk exposure, to demand more redistribution [@rehm_risks_2009]. In this framework, lower risk exposure driven by advantageous labour market positions held by higher classes is associated with lower redistributive preferences and higher perceived meritocracy.

In line with the previous argument, little evidence of the role of social class on meritocracy perception has been shown by the literature. For instance, @Smith2012 found that belonging to the emerging petty bourgeoisie is associated with higher levels of perceived meritocracy when compared to wage earners. These results align with the findings of @Sandoval2017, where skilled and service workers, compared to unskilled workers, attribute higher importance to effort to get ahead in life. However, a study on the relationship between social mobility and perceived meritocracy in Chile demonstrated that upward-mobile individuals towards the service class are less likely to consider that merit is rewarded [@Canales2015lse].

Social stratification research in Chile has shown regardless of the high-income inequality, social mobility is still high, mainly due to mobile individuals from lower-class origins to non-manual and service-class destinations [@Torche2005a]. However, regardless of the changes in the social structure, evidence shows that wages have remain relative stagnated and disposable income distribution has remained stable [@EspinozaNunez2014]. Additionally, during the last decades, the size of the middle class has grown driven by the increasing number of professional and technical occupations generated by tertiary education expansion [@Mclure_etal2014]. Nevertheless, this rise of human capital caused an increment in average wages but has not significantly altered market income inequality. In this line, a study on the cohort of individuals that joined the labour market from 1990 onwards, showed the expected positive returns of tertiary education on wages. Nevertheless, the variation in market income among the self-employed and service classes is high when taking into consideration the stratified access to tertiary education institutions with severe admission standards and their consequences on occupational prestige [@MacClure2012].

It can be argued that in contexts of high economic inequality, narratives of social mobility can be accompanied by frustrating income expectations or labour market discrimination experiences, which can lead to adverse perceptions toward meritocracy. For instance, a sense of frustration has been narrated by individuals with a lower-class background who have experienced upward social mobility [@barozet_clase_2011], pointing out that class background is a constant source of discrimination in the labour market [@cea_discriminacion_2020]. In this line, Chile is a country where the middle classes have grown significantly via upward social mobility. In addition, it has been shown that even among those who experienced improvements in their living conditions, this is not always correlated with positive perceptions about meritocracy and they can be even more critical due to the labour market constraints [@Canales2015lse]. This can be salient among individuals with higher educational attainment and those performing non-manual occupations, allowing us to expect that:

@@

The literature on subjective meritocracy has paid more attention to the structural position of individuals as an explanatory factor for perceptions of meritocracy [@Reynolds2014; @Duru-bellat2012; @Kunovich2007]. However, how individuals experience social inequality is affected by the status of the individuals' or groups [@garcia-castro_perception_2022;@castillo_perception_2022]. For example, perceived economic inequality in everyday life diminishes justification of inequality and increases demand for redistribution [@garcia-castro_perceiving_2020]. Nevertheless, it has been demonstrated that meritocracy primes reinforce negative evaluations of economically disadvantaged groups and strengthens internal attribution [@madeira_primes_2019].

Social interactions serve as anchors through which individuals rationalise their position in society. Perceived social status is influenced by normative appraisals of qualitative or ascribed status characteristics such as gender or quantitative like income or educational level [@Jasso2001]. This phenomenon has been conceptualised by the literature as subjective social status, understood as the perception of an individual or group's social standing in the social hierarchy in comparison to a reference group, such as society itself [@slomczynski1987; @Evans2004] or to their local communities or social classes [@Condon2020]. Measuring subjective social status is commonly done using the MacArthur Scale of Subjective Social Status, in which individuals report social status on a ladder that aims to represent the different strata of the social hierarchy [@adler_relationship_2000].

One of the main findings of this literature is the persistent bias toward the middle of the hierarchy when individuals are asked about their position in the social ladder, where high-status individuals tend to overestimate their position while underestimations are reported by lower-status individuals [@Lindemann2014]. In this regard, previous theoretical approaches have suggested that individuals are likely to compare themselves with people or groups perceived as similar. These reference groups contribute to diminishing cognitive dissonance and seek to assemble better estimations about their position in society [@festinger1954theory; @Merton1968].

The reference group and reality blend hypothesis (R&R) [@KelleyEvans1995] has suggested that perceptual biases can be explained through an availability heuristic, which reduces cognitive dissonance in situations with limited information [@Evans2004]. In this line, the characteristics of the closest reference groups shape subjective estimates of social standing, in which the images created about society are influenced by the homophily of their close networks [@Evans1992; @Evans2017]. For instance, @son_cognitive_2021 reveals that being a member of an organisation or practising a hobby contributes as symbolic pieces of information that shape the observer's cognitive patterns of homophily. In other words, people consider that individuals that share these attributes tend to be tied (e.g. friendship) and considered members of a similar group. In this regard, @kim_social_2021 have shown that individuals with close networks that concentrate on people with high-status occupations tend to report higher subjective social status in Chinese society, where social capital is considered essential. On the other hand, in an individualised society like Japan, the socioeconomic diversity of the close network does not influence subjective social status.

There is substantial evidence that socioeconomic status is a robust predictor of subjective social status, but it has been demonstrated that income, compared to education and class, has a stronger association with the perception of individuals about their standing in society [@Castillo2013;@IturraMellado2019]. Nevertheless, the literature has shown little evidence of the influence of subjective status on perceptions or preferences towards economic inequality. For example, @Vargas-Salfate2018 found that higher subjective status is related to a positive perception of the distributive system. Similarly, @Schneider2015 evidence a positive association of subjective status with the perceived justice income inequality and higher internal poverty attributions. In addition, @Castillo2018 found that higher subjective status is related to higher perceived meritocracy. Based on this evidence, our main prediction is:

@@

One of the claims of the R&R hypothesis is related to the interaction of both, subjective and material experiences [@Evans2017]. In this regard, the literature has demonstrated the positive influence status and class on subjective status. However, a less explored aspect is how subjective status can operate as a moderator of objective experiences on perceived meritocracy.

Studies on subjective social status have shown that among higher-income individuals, inconsistencies regarding perceived status tend to be higher [@Castillo2013]. For instance, @Andersson2018a has shown that disposable income, years of education and occupational prestige have a positive but non-linear relationship with subjective social status, indicating that individuals on the top rungs of the social hierarchy are those who also manifest the highest levels of variation in subjective status. On the one hand, attention to the role of status inconsistencies on individual preferences has focused on inconsistencies between income level and educational attainment [@Kreckel1993], demonstrating that a lower correlation between them can explain political preferences [@Lenski1967; @Varas2015]. On the other hand, studies like @Sosnaud2013 have analysed the inconsistencies between class identity and social class as an explanatory phenomenon of political preferences, establishing that those who tend to overestimate their class position prefer to vote for the conservatives. However, these results are not robust to socio-demographic controls.

The literature that considered the role of subjective social status as a moderating factor has mainly focused on how it is related to other subjective variables, paying less attention to how the perception of social status interacts with income or education. A study conducted by @Fatke2018 using data from the International Social Survey Programme (ISSP) analysed the role of diagrammatic perceptions of inequality in redistributive preferences and perceptions of meritocracy, focusing on the role of subjective status as a moderator of inequality perception. First, the results show that higher perceptions of inequality lead to higher preferences for redistribution and lower perceptions of meritocracy. Secondly, the relationship between inequality and meritocracy perception becomes weaker for those who perceive themselves as having a higher societal position. On the other hand, a study on European countries [@Bobzien2019] showed that economic inequality does not influence preferences for redistribution when controlled by perceived inequality because it negatively influences preferences for redistribution and is positively moderated by subjective status. In other words, when the subjective status is higher, the negative relationship of inequality perception with redistributive preferences tends to weaken. Another study by @bernardo_meritocracy_2021 analysed the relationship between intolerance of inequality and perceived meritocracy in a sample of bachelor students of high subjective social status, indicating that those who perceive themselves to have a higher subjective social status and perceive greater meritocracy tend to justify higher levels of economic inequality.

>>>>

---
 title: "Status, Class, Subjective status and meritocracy perception"
 # author: Julio Iturra ^[Bremen International Graduate School of Social Sciences (jiiturra@uni-bremen.de)]
 bibliography: "input/bib/ess-merit.bib"
 csl: "input/bib/apa6.csl"
 css: "input/css/custom.css"
 output:
     bookdown::pdf_document2:
         template: default
         toc: false
         number_sections: false
     bookdown::html_document2:
          number_sections: true
     bookdown::word_document2:
          toc: true
 geometry: margin=1in
 header-includes:
   # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel} # para decimales con commas en vez de punto
   - \usepackage{times}           # Times New Roman
   - \usepackage{caption}
   - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
   - \usepackage{graphicx}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
 linestretch: '1.5'
 fontsize: 12pt
 link-citations: yes
 linkcolor: blue
 editor_options:
   chunk_output_type: inline
 abstract: "What is the role of status, class and subjective status on perceived meritocracy? To what extent can subjective status affect the objective experiences given by income, education and social class on meritocracy perception? To answer these questions, we use nationally representative data from the 2018 wave of the Longitudinal Social Survey of Chile - ELSOC (N = 2,652). Perceived meritocracy constitutes the descriptive views about how effort and talent are rewarded in society. Contrary to our expectations, income and social class do not influence perceived meritocracy. However, in line with the hypothesis, individuals with higher educational attainment hold critical views about meritocracy, and those who allocate themselves close to the top groups in society perceive higher rewards regarding effort and talent. We also found that subjective status increases the meritocracy perception of those groups on top of the social hierarchy concerning income, education and social class."  
---

```{r render,include=FALSE, eval=FALSE}
 rmarkdown::render("paper-blind.Rmd")
```

@@

In recent decades, increasing economic inequality [@piketty_capital_2014] and stagnant social mobility [@oecd_broken_2018] have prompted social scientists to examine how societies perceive the functioning of meritocracy - the idea that effort and talent are rewarded - and its role in justifying economic inequality [@mijs_meritocracy_2020]. Meritocracy has been linked to justifying social inequalities, as individual achievements and class position are often attributed to individual agency rather than structural constraints [@frei_fruto_2020]. Studies have suggested that individuals' views of the labor market shape their perception of how economic resources are distributed [@Svallfors2006]. Belonging to non-manual and professional social classes is negatively associated with perceived meritocracy, especially in societies with high levels of economic inequality and low social mobility [@Canales2015lse].

Social stratification can be studied objectively, through the expression of social inequality in economic and cultural resources [@Kerbo2003], or subjectively, through individual experiences in the class structure [@Kluegel-Smith1986]. Class is often linked to occupation, labor market position, and levels of authority, while status refers to perceived superior, equal, or inferior relationships among individuals [@Chan2007]. Although there are scales of subjective social class [@Elbert2018], subjective social status has been more widely studied [@zhou_trajectory_2021]. Subjective social status refers to an individual's or group's perception of their standing in the social hierarchy compared to a reference group, such as society [@Evans2004] or their local community [@condon_economic_2020], and can influence perceptions of meritocracy [@Castillo2018]. People tend to compare themselves to similar groups to reduce cognitive dissonance and better estimate their societal position [@Merton1968;@festinger1954theory]. High-status individuals tend to underestimate their position while low-status individuals overestimate it [@Cruces2011]. The characteristics of an individual's reference groups can shape their subjective estimations and their image of society is influenced by the homophily of their close networks [@Evans2017]. Higher subjective social status is associated with higher perceived meritocracy and can also positively moderate the influence of class and status [@hvidberg_social_2020].

This study aims to understand the influence of objective class and status on meritocracy perception, as well as the role of subjective social status and its interaction with social class, education, and income.

@@

Merit is understood as the combination of effort and ability [@Sen2000]. Thus, the meritocratic ideal is focused on the individual, rather than structural factors in getting ahead in life [@mccall_exposure_2017].  In this regard, the measurement of meritocracy-related constructs has not been clear in distinguishing between descriptive and normative dimensions [@Castilloetal2019]. For instance, it is often to notice undifferentiated use of *Meritocratic Beliefs* [@mijs_how_2021a; @bernardo_meritocracy_2021] and *Perceived Meritocracy* [@Xian2017]. The lack of distinction can be problematic, especially when scrutinizing explanations on subjective views about meritocracy.

The concept of  _perception_ is understood as the subjective description of an object [@Aalberg2003]. In this regard, the descriptive (*how is*) dimension does not consider the normative evaluation of the object (*should be*). On the other hand, the normative dimension can be understood as preferences, influenced by previous beliefs and information affecting the evaluation of an object [@druckman_preference_2000]. In this regard, @Garcia2001 suggests perceived meritocracy is the degree to which individuals consider their society complies with the meritocratic principles. In other words, to what extent effort and talent are rewarded, indicating "how meritocracy is" rather than "how it should be".

@@

### Status

According to the rational self-interest approach [@Meltzer81], individuals with higher incomes perceive higher meritocracy because they accumulated economic resources through effort and talent. High status is also associated with lower inequality perception [@Evans2017] and higher justification of income inequality [@Trump2017].

There is evidence of a positive relationship between income and perceptions of meritocracy regarding the importance of individual merit in achieving success [@Reynolds2014; @Xian2017] and the extent to which intelligence and skills are rewarded [@Castilloetal2019; @Duru-bellat2012]. However, there is mixed evidence on how lower-income individuals rationalize inequality and meritocracy. Some research supports the "activated conflict theory," which suggests that the poor in unequal localities hold lower beliefs in meritocracy due to social comparisons that increase awareness of social status [@Newman2015]. Other research, based on the "relative power theory," challenges this finding, arguing that the rich hold more power due to income and wealth concentration, allowing them to disseminate legitimizing ideas about economic inequality, and the poor are less critical of dominant beliefs like meritocracy as a psychological mechanism for coping with deprivation [@Solt2016]. However, other research shows that low-income individuals in highly unequal contexts have firmer meritocracy beliefs, but the differences tend to fade as inequality declines [@morris_paradox_2022].

Based on the literature, we expect that individuals with higher economic resources will attribute more importance to merit and therefore perceive that rewards in society are distributed according to it. Therefore, we propose the following hypothesis:

> $H_{\text{1a}}$: Income is positively correlated with perceived meritocracy.

As income reflects an individual's economic resources, studies on social stratification have raised the importance of educational attainment on status achievement [@Goldthorpe2003]. In this line, research on meritocracy perception have discussed two theoretical approaches. On the one hand, the *reproductionist* approach [@Bourdieu2009; @bernstein_class_2003] supports the *socialisation* hypothesis, suggesting school socialisation operates as a normative reinforcement of meritocratic ideals, then educated individuals are more willing to support meritocracy as a legitimising mechanism [@Lampert2013]. On the other hand, the *instruction* hypothesis suggests perceived meritocracy tends to weaken in individuals with higher educational attainment because they would be more aware of the influence of structural factors on status achievement [@Duru-bellat2012] considering experiences of educational mobility [@Canales2015lse] and higher chances to engage with complex political ideas [@bubak_perceptions_2019].

On the one hand, evidence in favour of the socialisation hypothesis has pointed out education is associated with higher importance of individual factors in getting ahead in life [@Xian2017; @Reynolds2014; @mijs_paradox_2021], but also as a determinant of wages [@Kunovich2007]. On the other hand, evidence in favour of the instruction hypothesis shows that individuals more educated individuals disagree that people are rewarded for their effort and talent [@Duru-bellat2012; @Castilloetal2019]. Additionally, the instructional hypothesis can be complemented by the fact that greater awareness of inequality may experience feelings of injustice and relative deprivation since individuals with the same educational attainment have access to unequal rewards [@Adams1965; @Smith-eta2020_relative-deprivation]. In this sense, it has been shown higher educational attainment is associated with higher perceived inequality [@Castillo2012a_justice]. Thus, this evidence lead us to suggest:

> $H_{\text{1b}}$: education is negatively associated with perceived meritocracy.

@@

The relationship between structural position and meritocracy has mainly focused on income and educational attainment, setting social class aside from the discussion. However, occupational class not only represent the synthesis of educational achievement and income levels [@Ganzeboom1992] but also represents a symbol of power and authority in contemporary societies [@Chan2007]. Furthermore, studies on social mobility have shown the relevance of studying inequality through class schemes using occupation, autonomy and authority in the workspace as meaningful distinctions to categorise individuals [@Erikson1992], making them relatively homogeneous groups   [@Connelly2016]. However, it has been argued that common characteristics do not preclude class heterogeneity in terms of cultural consumption behaviour or political preferences [@Chan2004].

The literature on distributive preferences has suggested social class reflects individuals' experience in the labour market regarding job opportunities and wages [@Svallfors2006]. In a similar vein, @Kulin2013 demonstrated lower classes being exposed to higher labour market risks, such as the likelihood of job loss and lower wages, led individuals in material deprivation and risk exposure, to demand more redistribution [@rehm_risks_2009].

The literature has paid little attention to social class on meritocracy perception. For example, @Smith2012 found that individuals belonging to the petty bourgeoisie perceive higher levels of meritocracy compared to wage earners. These findings are supported by @Sandoval2017, which found that skilled and service workers attribute more importance to effort in achieving success compared to unskilled workers. However, @Canales2015lse found that upwardly mobile individuals in the service class perceived lower levels of meritocracy

Despite high income inequality, social mobility is still high in Chile, particularly for individuals from lower-class origins moving to non-manual and service-class destinations [@Torche2005a]. However, wages have remained stagnant and economic inequality has remained relatively stable [@EspinozaNunez2014]. This can lead to negative perceptions of meritocracy due to frustrating income expectations or experiences of discrimination in the labor market. For example, individuals with a lower-class background who have experienced upward mobility [@barozet_clase_2011] often report feeling frustrated by discrimination based on their class background [@cea_discriminacion_2020]. Upward mobility may not always result in satisfactory outcomes, which can further negative attitudes toward labor market constraints. Overall, we can hypothesis that:

@@

The literature on subjective meritocracy has paid more attention to the structural position of individuals as an explanatory factor for perceptions of meritocracy. However, how individuals experience social inequality is affected by the status of the individuals' or groups [@castillo_perception_2022]. For example, perceived economic inequality in everyday life diminishes justification of inequality and increases demand for redistribution [@garcia-castro_perceiving_2020]. Nevertheless, it has been demonstrated meritocracy primes reinforce negative evaluations of economically disadvantaged groups and strengthens internal attribution [@madeira_primes_2019].

Normative appraisals of qualitative or ascribed status characteristics such as gender or quantitative like income or educational level serve to explain status [@Jasso2001]. In this regard, social interactions serve as anchors through which individuals rationalise their societal position. The literature conceptualised it as subjective social status, understood as the perception of an individual or group's standing in the social hierarchy compared to a reference group, such as society itself [@slomczynski1987; @Evans2004] or local communities [@Condon2020]. A widely used measure is the MacArthur Scale of Subjective Social Status, in which individuals report social status on a ladder representing the strata of the social hierarchy [@adler_relationship_2000].

The literature has shown a persistent bias toward the middle of the hierarchy, where low-status individuals tend to overestimate their position while high-status individuals underestimate it [@Lindemann2014]. In this regard, previous theoretical approaches have suggested that individuals are likely to compare themselves with people or groups perceived as similar. These reference groups contribute to diminishing cognitive dissonance and seek to assemble better estimations about their position in society [@festinger1954theory; @Merton1968].

Social perception of meritocracy may be influenced by an individual's reference group and their perceived social standing within that group. The reference group and reality blend (R&R) hypothesis suggests that an individual's cognitive biases may be shaped by an availability heuristic, reducing cognitive dissonance in situations with limited information [@KelleyEvans1995; @Evans2004]. Close reference groups can shape an individual's subjective estimates of their social standing, and their imagined ideas about society may be influenced by the homophily within their social networks. Research has shown that individuals with networks consisting of people with high-status occupations have a higher subjective social status in Chinese society, where social capital is important [@kim_social_2021]. Membership in organizations or participation in hobbies may also shape cognitive patterns of homophily and influence an individual's perceived social standing [@son_cognitive_2021].

The link of subjective status with attitudes toward inequality has been little studied. For example, @Vargas-Salfate2018 found that higher subjective status is associated with higher system justification. Similarly, @Schneider2015 evidence a positive association of subjective status with the perceived justice income inequality and higher internal poverty attributions. In addition, @Castillo2018 found that higher subjective status is related to higher perceived meritocracy. Based on this evidence, our main prediction is:

@@

The R&R hypothesis suggests that both subjective and material experiences interact with each other [@Evans2017]. Studies have shown that status and class positively influence subjective status, but it is less understood how subjective status moderates the effect of objective experiences on perceived meritocracy.

Higher-income individuals tend to have higher inconsistencies in perceived status [@Castillo2013]. For example, income, education, and occupational prestige have a positive non-linear relationship with subjective social status, with higher status individuals showing more variation in subjective status [@Andersson2018a]. Inconsistencies between income level and educational attainment have been linked to political preferences [@Kreckel1993, @Lenski1967], and discrepancies between class identity and social class may also influence political preferences [@Sosnaud2013], but these results are not consistently supported by socio-demographic controls.

Studies on subjective social status as a moderating factor have primarily focused on its relationship to other subjective variables, rather than how it interacts with income or education. One study using data from the International Social Survey Programme (ISSP) found that higher perceptions of inequality lead to higher preferences for redistribution and lower perceptions of meritocracy, and that the relationship between inequality and meritocracy perception becomes weaker for those who perceive themselves as having a higher societal position [@Fatke2018]. Another study in European countries showed that economic inequality does not influence preferences for redistribution when controlled by perceived inequality, but is positively moderated by subjective status [@Bobzien2019]. A study on bachelor students with high subjective social status found that those who perceive themselves as having a higher status and greater meritocracy tend to justify higher levels of economic inequality [@bernardo_meritocracy_2021].

Experiments have shown that manipulating individuals' perceptions of their social status can alter their attitudes towards inequality. For example, in Argentina, people in the lowest and highest income deciles tend to have the most bias, with those in the lowest stratum overestimating their position and those in the highest underestimating it. Manipulating this bias can change attitudes towards redistribution [@Cruces2011]. In Sweden, older, educated people with higher cognitive skills tend to be more accurate about their position in the income distribution. It is believed that those with strong meritocratic beliefs will not be affected by informational interventions and will have fewer preferences for redistribution [@karadja_richer_2017]. Reference groups can also influence perceptions of distributive justice. Misperception is generally higher in large reference groups than in small ones. When informed of their actual position, those who overestimate their position around the median of any reference group, especially those in the lowest income stratum, tend to increase their perception of inequality and injustice [@hvidberg_social_2020].

Studies in Chile have found that three factors interact to shape perceptions of social status: socioeconomic status, economic and social capital, and moral perceptions of individual achievement [@macclure_naming_2020]. Individuals who identify as lower status tend to have low incomes, less education, and live in disadvantaged areas. However, some individuals who belong to middle or high socioeconomic strata may still identify as lower status due to living in affluent neighborhoods or having higher educational credentials, which symbolize higher social and cultural capital [@Mclure-etal2019, p. 18]. Those who identify as high status tend to consistently have higher socioeconomic status, with educational meritocracy often serving as a criterion for their social position [@Mclure-etal2019,p.19].

Subjective social status can reflect an individual's position in the social structure as well as their experience of economic inequality through comparisons with nearby reference groups. Therefore, we predict that differences in subjective status can lead to differences in perceived meritocracy within socio-economic groups. Thus, we predict that:
